---
title: "Comparing Students' Grades"
author: "Jay Ralyea, William Cull, and John Hope"
date: "5/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # whole bunch of useful packages
library(tidymodels)
library(modelr)
library(randomForest) # builds random forest models
library(caret) # builds models and partitions data
library(viridis)
```



```{r}
## Load in the data
math <- read_csv("student-mat.csv")
por <- read_csv("student-por.csv")

## Data exploration
quantile(math$G3)
quantile(por$G3)

## Percent zeros
table(math$G3)[1] / nrow(math)
table(por$G3)[1] / nrow(por)

## Factor variables in each dataset
factors <- c("school", "sex", "address", "famsize", "Pstatus", "Medu",
             "Fedu", "Mjob", "Fjob", "reason", "guardian", "traveltime",
             "studytime", "failures", "schoolsup", "famsup", "paid",
             "activities", "nursery", "higher", "internet", "romantic")

math[factors] <- sapply(math[factors], as.factor)
por[factors] <- sapply(por[factors], as.factor)

## Remove G1 and G2 grades (determine G3 grade)
math <- math %>%
  select(-G1, -G2)
por <- por %>%
  select(-G1, -G2)

## Change the G3 score to be poor, good, or excellent
math$G3 <- cut(math$G3,
                   breaks = c(-Inf, 8, 13, Inf),
                   labels = c("poor", "good", "excellent"))
por$G3 <- cut(por$G3,
                  breaks = c(-Inf, 10, 13, Inf),
                  labels = c("poor", "good", "excellent"))
  
```


# EDA
## Data Issues
```{r}
# School
math %>%
  ggplot(aes(x = school, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(title = "Distirbution of Math G3 Scores by School", 
       x = "School", y = "Count of Scores", fill = "G3 Score") +
  scale_x_discrete(labels = c("GP" = "Gabriel Pereira", 
                              "MS" = "Mouzinho da Silveira"))

por %>%
  ggplot(aes(x = school, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(title = "Distirbution of Portuguese G3 Scores by School", 
       x = "School", y = "Count of Scores", fill = "G3 Score") +
  scale_x_discrete(labels = c("GP" = "Gabriel Pereira", 
                              "MS" = "Mouzinho da Silveira"))
```

```{r}
# Gender
math %>%
  ggplot(aes(x = G3, fill = G3)) + 
  geom_bar() + facet_grid(cols = vars(sex)) +
  scale_fill_viridis_d() +
  theme_bw()

por %>%
  ggplot(aes(x = G3, fill = G3)) + 
  geom_bar() + facet_grid(cols = vars(sex)) +
  scale_fill_viridis_d() +
  theme_bw()
```

```{r}
# Family size
math %>%
  ggplot(aes(x = famsize, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(title = "Distirbution of Math G3 Scores by Family Size",
       y = "Count of Scores", x = "Family Size", fill = "G3 Score") +
  scale_x_discrete(labels = c("GT3" = "Greater than 3",
                              "LE3" = "Less than or equal to 3"))

por %>%
  ggplot(aes(x = famsize, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(title = "Distirbution of Portuguese G3 Scores by Family Size",
       y = "Count of Scores", x = "Family Size", fill = "G3 Score") +
  scale_x_discrete(labels = c("GT3" = "Greater than 3",
                              "LE3" = "Less than or equal to 3"))
```


```{r}
# Parental job - Math
math %>%
  ggplot(aes(x = Mjob, fill = G3)) + 
  geom_bar(position = "dodge") + 
  scale_fill_viridis_d() +
  labs(title = "Distribution of Math G3 Scores by Mother's Job",
       x = "Mother's Job", y = "Count of Score", fill = "G3 Score") +
  theme_bw()

math %>%
  ggplot(aes(x = Fjob, fill = G3)) + 
  geom_bar(position = "dodge") + 
  scale_fill_viridis_d() +
  labs(title = "Distribution of Math G3 Scores by Father's Job",
       x = "Father's Job", y = "Count of Score", fill = "G3 Score") +
  theme_bw()
```


```{r}
# Parental job - Portuguese
por %>%
  ggplot(aes(x = Mjob, fill = G3)) + 
  geom_bar(position = "dodge") + 
  scale_fill_viridis_d() +
  labs(title = "Distribution of Portuguese G3 Scores by Mother's Job",
       x = "Mother's Job", y = "Count of Score", fill = "G3 Score") +
  theme_bw()

por %>%
  ggplot(aes(x = Fjob, fill = G3)) + 
  geom_bar(position = "dodge") + 
  scale_fill_viridis_d() +
  labs(title = "Distribution of Portuguese G3 Scores by Father's Job",
       x = "Father's Job", y = "Count of Score", fill = "G3 Score") +
  theme_bw()
```

## Potentially Important Factors

```{r}
# Failures
math %>%
  ggplot(aes(x = failures, fill = G3)) + 
  geom_bar(position = "dodge") + 
  scale_fill_viridis_d() +
  labs(title = "Distribution of Math G3 Scores by Number of Failures",
       x = "Number of Failures", y = "Count of Scores", fill = "G3 Score") +
  theme_bw()

por %>%
  ggplot(aes(x = failures, fill = G3)) + 
  geom_bar(position = "dodge") + 
  scale_fill_viridis_d() +
  labs(title = "Distribution of Portuguese G3 Scores by Number of Failures",
       x = "Number of Failures", y = "Count of Scores", fill = "G3 Score") +
  theme_bw()
```

```{r}
# Age
math %>%
  ggplot(aes(x = age, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() + 
  theme_bw() +
  scale_x_continuous(breaks = seq(15, 22, 1)) +
  theme(panel.grid.minor.x = element_blank())

por %>%
  ggplot(aes(x = age, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() + 
  theme_bw() +
  scale_x_continuous(breaks = seq(15, 22, 1)) +
  theme(panel.grid.minor.x = element_blank())
```

Generally, the older the student is the more likely they are to succeed. This is especially true for their Portuguese class. The exception to this rule is this generalization is students who are over 18 years of age. Students 19 years old or older likely have extenuating circumstances that are keeping them in the secondary school. Those students may have been held back in the past due to performance or behavioral issues. Either of these situations is likely to have a negative effect on the student's performance in the classroom.

```{r}
# Family size
math %>%
  ggplot(aes(x = famsize, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(title = "Distirbution of Math G3 Scores by Family Size",
       y = "Count of Scores", x = "Family Size", fill = "G3 Score") +
  scale_x_discrete(labels = c("GT3" = "Greater than 3",
                              "LE3" = "Less than or equal to 3"))

por %>%
  ggplot(aes(x = famsize, fill = G3)) + 
  geom_bar(position = "dodge") +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(title = "Distirbution of Portuguese G3 Scores by Family Size",
       y = "Count of Scores", x = "Family Size", fill = "G3 Score") +
  scale_x_discrete(labels = c("GT3" = "Greater than 3",
                              "LE3" = "Less than or equal to 3"))
```

 

```{r}
## Split the data for model building

## Create the partition
set.seed(2000) # ensure reproducibility
math.part <- createDataPartition(math$G3, times=1, p=0.8, list=FALSE)
set.seed(2000) # ensure reproducibility
por.part <- createDataPartition(por$G3, times=1, p=0.8, list = FALSE)

## Math training and testing data
math.training <- math[math.part,]
math.test <- math[-math.part,]

## Portuguese training and testing data
por.training <- por[por.part,]
por.test <- por[-por.part,]
```




```{r}
## Divide data: training/validation/testing
math.div <- math %>%
  initial_split(prop = 0.6, strata = school)

math.div2 <- math.div %>%
  testing() %>%
  initial_split(prop = 0.5, strata = school)

math.train <- training(math.div)
math.validate <- training(math.div2)
math.test <- testing(math.div2)


#general rule to start with the mytry value is square root of the predictors
mytry_tune <- function(x){
  xx <- dim(x)[2]-1
  sqrt(xx)
}

mytry_tune(math)


## Build random forest
## Four variables are eligible for a split at each node
RF.mod1 <- randomForest(G3 ~ ., data = math.train,
                        mtry = 6, importance = TRUE)

RF.mod1

RF.mod1 %>%
  importance()
```































